# Reproducible Research: Peer Assessment 1
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data

This assignment requires activity.zip file which can be downloaded from [this location](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  
We first unzip the file and extract the activity.csv file.  
Then let us take a look at the structure of the csv by using the head function and the summary function.

```{r}
unzip("activity.zip");
activity<-read.csv("activity.csv");
head(activity);
summary(activity);
activity$date <- as.Date(activity$date);
```


## What is mean total number of steps taken per day?
For this part of the exercise, we will first sum the number of steps by date omitting the NAs.  
_Please note that NAs are automatically removed by using aggregate function in this form._  
```{r}
activityByDate <- aggregate(steps ~ date, data = activity, sum);
```
First, we will make a histogram of the total number of steps taken each day.  
```{r}
hist(activityByDate$steps, col="red", main="Histogram of total steps taken daily", xlab="Total steps taken daily");
```
Now we will take a mean and median of the resulting data.
```{r}
activityMean <- mean(activityByDate$steps);
activityMedian <- median(activityByDate$steps);
```

Mean of total number of steps taken each day is **`r activityMean` steps**.  
Median of total number of steps taken each day is **`r activityMedian` steps**.

## What is the average daily activity pattern?
For this question, we will first take mean of steps by interval.  
Then we will use the *qplot* from *ggplot2* library to draw a *linegraph* of mean steps by interval.  
```{r}
activityByInterval <- aggregate(steps ~ interval, data = activity, mean);
library(ggplot2);
qplot(interval, steps, data=activityByInterval, geom=c("line"), main="Mean daily steps by interval", ylab="Mean Steps");
```

Next, we will use the *which* function to find out which daily interval contains the maximum of mean steps.

```{r}
maxSteps <- max(activityByInterval$steps);
maxStepsInterval <- activityByInterval[which(activityByInterval$steps == maxSteps), 1];
```

The daily interval containing maximum number of mean steps (**`r maxSteps` steps**) is **Interval  `r maxStepsInterval`**.

## Imputing missing values
First we will find all missing values using *which* function.  

```{r}
missingIndex <- which(is.na(activity$steps));
missingCount <- nrow(as.array(missingIndex));
```

**We find that `r missingCount` values are missing.**  
  
Next we will replace those missing values by the mean steps of that interval. We have these already calculated in the **activityByInterval**.  
  
Lets call the new data frame as filledActivity.

```{r}
filledActivity <- activity;
for(i in missingIndex){
        filledActivity$steps[i] <- activityByInterval$steps[activityByInterval$interval == filledActivity$interval[i]];
}
```

Now, we will redo the following  
1. Aggregation by day  
2. Histogram by day  
3. Mean and Median  

```{r}
filledActivityByDate <- aggregate(steps ~ date, data = filledActivity, sum);
hist(filledActivityByDate$steps, col="red", main="Histogram of total steps taken daily (filled)", xlab="Total steps taken daily (filled)");
filledActivityMean <- mean(filledActivityByDate$steps);
filledActivityMedian <- median(filledActivityByDate$steps);
```

New Mean of total number of steps taken each day is **`r filledActivityMean` steps** vs Old Mean **`r activityMean` steps**.  
New Median of total number of steps taken each day is **`r filledActivityMedian` steps** vs Old Median **`r activityMean` steps**.  
**There is no material difference in mean and median of the steps taken per day.** This is because the missing data was missing for the full days. And when the data was filled with means across days, there was no impact. This can be verified by the following scatterplot for date-interval patterns in missing values.  
```{r}
qplot(date, interval, data=activity[missingIndex, ])
```

## Are there differences in activity patterns between weekdays and weekends?


```{r}
filledActivity$day <- weekdays(filledActivity$date);
#qplot(interval, steps, data=filledActivity, facets=steps ~ day);
```
